---
title: "Gov 50 Final Project"
author: "Jenny Lu"
description: "Exploring the effects of the pandemic on public transport systems"
output:
  distill::distill_article:
    self_contained: false
---

## Introduction
As we approach four years since the start of the COVID-19 pandemic, our world seems to have returned to normalcy – or, at least, a new normal. There is no question that the pandemic touched multiple facets of our lives, from the way in which we learn, to the forms of interactions that maintain our interpersonal relationships. However, lingering effects of the pandemic still remain, like the rise of remote/hybrid work models even as many have returned to the office. One such lingering effect is seen in public transportation. Public transportation, especially in large cities, is the way in which thousands of people interact and cross paths with each other on a daily basis, so it’s natural that ridership levels took a hit during the pandemic. However, in the time since, has ridership recovered? 

In this analysis, I’m interested in exploring the extent to which public transport has been impacted by the COVID-19 pandemic. More specifically, which form of public transport has been hit the hardest? I plan to examine data from the New York Metropolitan Transportation Authority (MTA), containing statistics on ridership and the corresponding proportion of pre-pandemic ridership levels for each of the forms of public transport offered under the MTA in the greater New York City metropolitan area. I hypothesize that the Subway form of public transport suffered the largest impact on ridership from the pandemic, due to it having the highest ridership levels relative to all other forms of public transport, as well as the unavoidable proximity riders experience, and as a consequence, is likely viewed to be of highest-risk.

## Data
The MTA is a public benefit corporation responsible for public transportation in the state of New York. This dataset provides systemwide ridership and traffic estimates for subways (including the Staten Island Railway), buses, Long Island Rail Road, Metro-North Railroad, Access-A-Ride, and Bridges and Tunnels, beginning March 1st, 2020 (April 1st, 2020 for LIRR and Metro-North), and provides a percentage comparison against a comparable pre-pandemic date (the equivalent month in 2019). Ridership figures (and subsequently, the `% of Comparable Pre-Pandemic Day` values [the dependent variable in this analysis]) are determined through what can be generalized as a census method -- a combination of MetroCard, OMNY swipes/taps, ticket sales, scheduled trips, and toll collection systems -- tracked over time (the independent variable) on a daily basis from March 1st, 2020 to present.

This analysis follows a longitudinal study method, as we are observing the same sample (ridership levels for each of the forms of transport) over an extended period of time (2020 to present) and are collecting this data without direct "treatment". However, the "treatment" in this analysis can be seen as being the pandemic, the "treatment group" being the ridership levels pre-pandemic (2019 data), and the "control group" being the ridership levels after the onset of the pandemic (post-March 1st, 2020). Through the longitudinal research design, we are able to measure the effect the "treatment" (the pandemic) had on ridership over time. Below is a boxplot summarizing the `% of Comparable Pre-Pandemic Day` values for each type of transportation. As seen in the following boxplot, the median `% of Comparable Pre-Pandemic Day` value hovers around 50-60% for most forms of transport, with some notably high exceptions for Bridges and Tunnels and Access-a-Ride.

### Tidying the Data
```{r setup}
options(width = 100)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(infer)
```

```{r}
ridership <- read_csv("data/ridership.csv")

ridership_df <- as.data.frame(t(ridership))

names(ridership_df) <- ridership_df[1,]

ridership_df <- ridership_df[c(-1,-2,-4,-6,-8,-10,-12,-14),]

ridership_df$type <- rownames(ridership_df)

ridership_prop <- ridership_df |>
  pivot_longer(!type, names_to = "date", values_to = "percentage")

ridership_prop$type <- gsub(
  ": % of Comparable Pre-Pandemic Day", 
  "", 
  as.character(ridership_prop$type))

ridership_prop$percentage <- as.numeric(ridership_prop$percentage)

ridership_prop$date <- as.Date(ridership_prop$date,"%m/%d/%Y")

ridership_prop <- ridership_prop |>
  na.omit()
```

### Summary Boxplot of Dependent Variable
``` {r}
ridership_prop |>
  ggplot(aes(x = type, y = percentage)) +
  geom_boxplot() +
  labs(
    x = "Type of Transport",
    y = "% of Comparable Pre-Pandemic Day",
    title = "% of Comparable Pre-Pandemic Day Values, by Type of Transport"
  ) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5))

```

## Results
Below is a scatterplot graph with lines of best fit (individual points removed to maintain the graph's readability) depicting `% of Comparable Pre-Pandemic Day` values from March 1st, 2020 to present for each form of transportation. Lines should each start at a y-value of 1, but because measurements began on March 1st, 2020 and COVID restrictions were enacted soon after the start of data collection, the `geom_smooth()` function causes the `% of Comparable Pre-Pandemic Day` values to appear to start at far less than 1.

As seen in the graph, buses were initially most severely impacted by the pandemic, with their `% of Comparable Pre-Pandemic Day` measurements dropping the most relative to any other form of transportation. Other forms of transport, such as the subway, LIRR, Metro-North, and Staten Island Railway, also had significant drops in ridership relative to a comparable pre-pandemic day. However, Bridges and Tunnels and Access-a-Ride had relative increases in ridership, which make sense, given individuals worried about transmission tended to prefer forms of transport that are not as close-contact as subways or buses. Following the `% of Comparable Pre-Pandemic Day` measurements for each of the forms of transport over time, we can see that Bridges and Tunnels and Access-a-Ride ridership levels have recovered to near pre-pandemic levels; however, other forms of public transport such as subways, buses, and the Staten Island Railway are still suffering the effects of the pandemic with ridership levels hovering at around 60% of pre-pandemic levels.

### Main Relationship of Interest Plot
```{r}
ridership_prop |>
  group_by(type) |>
  ggplot() + 
  geom_smooth(aes(date, percentage, group = type, color = type)) +
  labs(
    x = "Date",
    y = "% of Comparable Pre-Pandemic Day",
    title = "% of Comparable Pre-Pandemic Day Values, by Type of Transport",
    col = "Type of Transport"
  )
```

### Regression
Below is a regression analysis of the relationship between time and `% of Comparable Pre-Pandemic Day` values, conditional on the type of transportation. This regression has no singular main coefficient of interest; instead, each type of transportation is interpreted individually. The regression indicates that post-March 1st, 2020 ridership on most forms of transportation saw a decrease in their ridership levels relative to their respective pre-pandemic ridership (decrease of 0.233 for buses, 0.254 for LIRR, 0.330 for Metro-North, 0.418 for Staten Island Railway, and 0.256 for subways), with the Staten Island Railway having the greatest decrease. However, other forms of transportation saw an increase in ridership; for example, Bridges and Tunnels experienced an increase in ridership levels of 0.147 relative to its pre-pandemic ridership. Each of these types of transportation have p-values of less than 0.001, so assuming an alpha value of 0.05, their respective estimated coefficients are statistically significant, and I believe the coefficients represent a causal effect.

```{r}
ate <- lm(
  percentage ~ date + type,
  data = ridership_prop
)
summary(ate)

var_labels <- c(
  "(Intercept)" = "Intercept",
  "date" = "Date (starting March 1st, 2020)",
  "typeAccess-a-Ride" = "Access-a-Ride",
  "typeBridges and Tunnels" = "Bridges and Tunnels",
  "typeBuses" = "Buses",
  "typeLIRR" = "LIRR",
  "typeMetro-North" = "Metro-North",
  "typeStaten Island Railway" = "Staten Island Railway",
  "typeSubways" = "Subways"
)
modelsummary::modelsummary(ate,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           coef_map = var_labels,
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```

## Conclusion
From both the scatterplot graph of best-fit lines and the regression, it's clear that the different types of transportation were affected differently by the pandemic; some had increases in ridership levels relative to pre-pandemic levels, while most had decreases. My original hypothesis was that the subway form of transportation suffered the largest negative impact from the pandemic; however, the analyses performed indicate that the Staten Island Railway was actually the form of transportation that, to date, has the lowest percent ridership relative to pre-pandemic levels. There are, however, limitations to my analysis in this study: possible confounding factors exist, such as the disparity in return-to-office policies between white- and blue-collar jobs. As mentioned in the introduction, many white-collar jobs adapted to be work-from-home and have remained so. As a result, those who used to take public transportation may no longer need to, or as frequently. In this case, health factors and pandemic-related restrictions/precautionary behaviors are not the sole cause of this decline in ridership, but rather, other more indirect factors such as these hybrid work policies are affecting ridership. Another limitation is this dataset's lack of data pre-March 1st, 2020, which makes it difficult to understand the relative gravity of these ridership declines -- have there been other large historical events in the past that affected public transportation ridership? And to what degree? This analysis could be improved with this additional data on historical trends, as well as with more specific data on ridership demographics to create a more accurate picture of how certain groups have been affected by the pandemic.
